---
title: "Riley_Project1"
format: html
---

# Project Description
This project will manipulate and process data sets from a census API and then summarize and plot returned data, mainly using custom functions.

# Author Information
Author: Shannon Riley
Organization: NCSU ST 558
Date: Fall 2026 (Completed ___)

## Data
This project will utilize the Public Use Microdata Sample (PUMS) Census API. We will consider the years 2021-2024 for the basic 1-Year Estimates (will not include Puerto Rico).

```{r}
#| echo: false
#| output: false
# Load required packages
library(dplyr)     # to perform basic EDA
library(ggplot2)   # to make nice plots
library(hms)       # to parse times
library(httr)      # to query an API
library(jsonlite)  # to parse results of API query, tidyjson is another possible package to use
library(lubridate) # to parse times
library(readr)     # to read in CSV data
library(stringr)   # to create clean column names
library(tidyr)     # to clean up tibbles
```

First, the JSON files with the variable descriptions for each year (2021-2024) will be downloaded. We only want to parse the following subset of variables:
+ AGEP - Age
+ GASP - Gas cost (monthly cost, use ADJHSG to adjust GASP to constant dollars)
+ GRPIP - Gross rent as a percentage of household income past 12 months
+ JWAP - Time of arrival at work (hour and minute)
+ JWDP - Time of departure for work (hour and minute)
+ JWMNP - Travel time to work
+ PWGTP - PUMS person weight
+ FER - Gave birth to child within the past 12 months
+ HHL - Household language
+ SCH - School enrollment
+ SCHL - Educational attainment
+ SEX - Sex
+ ST (for 2021, 2022)/STATE (for 2023, 2024) - State Code based on 2020 Census definitions
+ REGION - Region code based on 2020 Census definitions
+ DIVISION - Division code based on 2020 Census definitions

To avoid repetitive code, write a function to pull each year's variable and their values and put them in a list, then save the all the lists to an *.rds* file:
/
First, create a vector the years we're interested in:
```{r}
years <- c('2021', '2022', '2023', '2024')
```

Next, create two vectors of the desired variables, since one the "state" variable is named differently in different years:
```{r}
subVar2122 <- c('AGEP', 'GASP', 'GRPIP', 'JWAP', 'JWDP', 'JWMNP', 'PWGTP', 'FER', 'HHL', 'SCH', 'SCHL', 'SEX', 'ST', 'REGION', 'DIVISION')
subVar2324 <- c('AGEP', 'GASP', 'GRPIP', 'JWAP', 'JWDP', 'JWMNP', 'PWGTP', 'FER', 'HHL', 'SCH', 'SCHL', 'SEX', 'STATE', 'REGION', 'DIVISION')
```

Now, write a function to access the JSON, parse the data, and pull out the desired variables and their values for a given year:
```{r}
getVarVal <- function(year) {
  # Download the JSON
  getURL <- GET(paste0('https://api.census.gov/data/', year, '/acs/acs1/pums/variables.json'))
  # Parse the JSON
  jsonParse <- fromJSON(rawToChar(getURL$content))
  # Create an empty list for the variables and their values to go in
  subVarLs <- list()
  # Check what the value of year is and use the first for loop if it is 2021 or 2022 and the second for loop if it is 2023 or 2024
  if (year == '2021' | year == '2022') {
    # Loop through each value in the subset of variables vector
    for (i in subVar2122) {
      # Pull the variable and its values out of the large list and put them in a list
      subVarLs <- c(subVarLs, jsonParse[["variables"]][i])
    }
  } else {
    for (i in subVar2324) {
      subVarLs <- c(subVarLs, jsonParse[["variables"]][i])
    }
  }
  # Return the list with the variables and their values
  return(subVarLs)
}
```

Run this function on each value in the vector of years:
*For some reason this doesn't work if I put assign() in return(), but does otherwise
```{r}
for (year in years) {
  assign(x = paste0('varList', year), value = getVarVal(year))
}
```

Now save all four year lists to an .rds file
```{r}
write_rds(list(varList2021 = varList2021, varList2022 = varList2022, varList2023 = varList2023, varList2024 = varList2024), file = 'data/varLists.rds', compress = 'none')
```


<!-- Pull each year's list variable values using the *GET()* function from the *httr* package: -->
<!-- ```{r} -->
<!-- varVal21 <- GET('https://api.census.gov/data/2021/acs/acs1/pums/variables.json') -->
<!-- str(varVal21, max.level = 1) -->
<!-- ``` -->

<!-- Parse the JSON with the variables and the values each variable can take then pull out the desired variables (listed above): -->
<!-- ```{r} -->
<!-- varParse <- fromJSON(rawToChar(varVal21$content)) -->
<!-- subVar <- c('AGEP', 'GASP', 'GRPIP', 'JWAP', 'JWDP', 'JWMNP', 'PWGTP', 'FER', 'HHL', 'SCH', 'SCHL', 'SEX', 'ST', 'REGION', 'DIVISION') -->

<!-- subVarLs <- list() -->
<!-- for (i in subVar) { -->
<!--   subVarLs <- c(subVarLs, varParse[["variables"]][i]) -->
<!-- } -->

<!-- ``` -->



Create a correctly formatted data frame of JWAP values to parse the JWAP variable if the user requests it:
```{r}
# Pull out the named values possible for the JWAP variable and convert them to a one-column data frame
  jwapValues <- t(data.frame(varList2024$JWAP$values$item))
  # Move the names of the values into a column
  jwapValues <- data.frame(item = as.numeric(names(varList2024$JWAP$values$item)), 
                           value = jwapValues, 
                           row.names = NULL)
  # Remove the observation for NA values
  jwapValues <- subset(jwapValues, item != 0)
  # Split the value column into two columns in order to calculate the midpoint of the time period
  jwapValues <- separate_wider_delim(jwapValues, 
                                     cols = 'value', 
                                     delim = ' to ',
                                     names = c('bRange', 'eRange'))
    # Replace 'a.m.' and 'p.m.' with 'AM' and 'PM' for easier parsing to time class
  jwapValues$bRange <- str_replace_all(jwapValues$bRange, c('a.m.' = 'AM', 'p.m.' = 'PM'))
  jwapValues$eRange <- str_replace_all(jwapValues$eRange, c('a.m.' = 'AM', 'p.m.' = 'PM'))
  # Convert the 'bRange' and 'eRange' variables into time objects
  jwapValues$bRange <- parse_date_time(jwapValues$bRange, '%I:%M %p')
  jwapValues$eRange <- parse_date_time(jwapValues$eRange, '%I:%M %p')
  jwapValues$midpoint <- as_hms(jwapValues$bRange + ((jwapValues$eRange - jwapValues$bRange)/2))
  # Add a value for missing time values back in (to correspond with the original value "N/A (not a worker; worker who worked from home)")
  jwapValues <- rbind(jwapValues, c(0, NA, NA, NA))
```

Create a correctly formatted data frame of JWDP values to parse the JWDP variable if the user requests it:
```{r}
# Pull out the named values possible for the JWDP variable and convert them to a one-column data frame
  jwdpValues <- t(data.frame(varList2024$JWDP$values$item))
  # Move the names of the values into a column
  jwdpValues <- data.frame(item = as.numeric(names(varList2024$JWDP$values$item)), 
                           value = jwdpValues, 
                           row.names = NULL)
  # Remove the observation for NA values
  jwdpValues <- subset(jwdpValues, item != 0)
  # Split the value column into two columns in order to calculate the midpoint of the time period
  jwdpValues <- separate_wider_delim(jwdpValues, 
                                     cols = 'value', 
                                     delim = ' to ',
                                     names = c('bRange', 'eRange'))
    # Replace 'a.m.' and 'p.m.' with 'AM' and 'PM' for easier parsing to time class
  jwdpValues$bRange <- str_replace_all(jwdpValues$bRange, c('a.m.' = 'AM', 'p.m.' = 'PM'))
  jwdpValues$eRange <- str_replace_all(jwdpValues$eRange, c('a.m.' = 'AM', 'p.m.' = 'PM'))
  # Convert the 'bRange' and 'eRange' variables into time objects
  jwdpValues$bRange <- parse_date_time(jwdpValues$bRange, '%I:%M %p')
  jwdpValues$eRange <- parse_date_time(jwdpValues$eRange, '%I:%M %p')
  jwdpValues$midpoint <- as_hms(jwdpValues$bRange + ((jwdpValues$eRange - jwdpValues$bRange)/2))
  # Add a value for missing time values back in (to correspond with the original value "N/A (not a worker; worker who worked from home)")
  jwdpValues <- rbind(jwdpValues, c(0, NA, NA, NA))
```

Create a vector of educational attainment values to parse the SCHL variable if the user requests it:
```{r}
schlLevels <- c(names(varList2024[["SCHL"]][["values"]][["item"]]))
schlLabels <- c(unname(unlist(varList2024[["SCHL"]][["values"]][["item"]])))
```



Now write a function allow users to easily query the PUMS API:

First, get a given URL
```{r}
pums2021 <- GET('https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=533d84cb67d9d71606e0baf0f4e7f32e4af7c4ce')
str(pums2021, max.level = 1)
```

Then, take the result of GET() and turn it into a nice tibble
```{r}
pumstbl <- as_tibble(fromJSON(rawToChar(pums2021$content)))
colnames(pumstbl) <- c(pumstbl[1,])
pumstbl <- slice(pumstbl, -1)
pumstbl
```

Create vectors of allowable values for *specificGeography* argument in *getPUMS()* function for user to reference if necessary:
```{r}
regionValues <- unname(unlist(c(varList2021$REGION$values)))
divisionValues <- unname(unlist(c(varList2021$DIVISION$values)))
stateValues <- unname(unlist(c(varList2021$ST$values)))
```


Combine these functions with input specifications to make a single function:
```{r}
getPUMS <- function(year = '2024', numericVariables = c('AGEP', 'PWGTP'), categoricalVariables = c('SEX'), geographyLevel = 'STATE', specificGeography = NULL, APIkey, ...) {
  
  # Read in .rds file with lists of variables and their values for each year
  varLists <- readRDS('data/varLists.rds')
  
  
  # Vectors to check inputs against
  allowNumVar <- c('PWGTP', 'AGEP', 'GASP', 'GRPIP', 'JWAP', 'JWDP', 'JWMNP')
  allowCatVar <- c('FER', 'HHL', 'SCH', 'SCHL', 'SEX')
  allowGeoLev <- c('REGION', 'DIVISION', 'STATE')
  allowRegion <- unname(unlist(c(varLists$varList2021$REGION$values)))
  allowDivision <- unname(unlist(c(varLists$varList2021$DIVISION$values)))
  allowState <- unname(unlist(c(varLists$varList2021$ST$values)))
  
  
  # Check inputs and return error message if criteria aren't met
  if(year != '2021' | year != '2022' | year != '2023' | year != '2024') {
    stop('"year" must be a single year between 2021 and 2024, input as a character string.')
  }
  if(!('PWGTP' %in% numericVariables) | length(numericVariables) < 2 | !(numericVariables %in% allowNumVar)) {
    stop('numericVariables must be a vector containing PWGTP and at least one of the following variables: AGEP, GASP, GRPIP, JWAP, JWDP, and JWMNP.')
  }
  if(length(categoricalVariables) < 1 | !(categoricalVariables %in% allowCatVar)) {
    stop('categoricalVariables must be a vector containing at least one of the following variables: FER, HHL, SCH, SCHL, and SEX.')
  }
  if(length(geographyLevel) > 1 | !(geographyLevel %in% allowGeoLev)) {
    stop('geographyLevel must be only one of the following variables: REGION, DIVISION, or STATE.')
  }
  if(specificGeography != 'ALL' | specificGeography != 'all' | specificGeography != 'All' | geographyLevel == 'REGION' & !(specificGeography %in% allowRegion) | geographyLevel == 'DIVISION' & !(specificGeography %in% allowDivision) | geographyLevel == 'STATE' & !(specificGeography %in% allowState)) {
  stop('specificGeography must be ALL or one of the allowable values associated with the specified value of geographyLevel. See vectors regionValues, divisionValues, and stateValues for allowable values for specificGeography.')
  }
  if(!is.character(API_key)) {
    stop('API key must be character string.')
  }
  
  
  # Specify default values for 'specificGeography' if user leaves argument NULL
  if (is.null(specificGeography) & geographyLevel == 'REGION') {
    specificGeography <- 'West'
  } else if (is.null(specificGeography) & geographyLevel == 'DIVISION') {
    specificGeography <- 'Pacific (West region)'
  } else if (is.null(specificGeography) & geographyLevel == 'STATE') {
    specificGeography <- 'Oregon/OR'
  }
  
  # Get numeric values corresponding to specificGeography inputs to paste into
  # URL so that the user can put in a 'expected' value rather than a number
  if (geographyLevel == 'REGION') {
    specificGeographyNum <- names(which(unlist(varLists$varList2021$REGION$values$item) == specificGeography))
  } else if (geographyLevel == 'DIVISION') {
    specificGeographyNum <- names(which(unlist(varLists$varList2021$DIVISION$values$item) == specificGeography))
  } else if (geographyLevel == 'STATE') {
    specificGeographyNum <- names(which(unlist(varLists$varList2021$ST$values$item) == specificGeography))
  }
  
  
  # Get data from API
  pums <- GET(paste0('https://api.census.gov/data/', year, '/acs/acs1/pums?get=', paste(numericVariables, collapse = ','), ',', paste(categoricalVariables, collapse = ','), '&for=', geographyLevel, ':', specificGeographyNum, '&key=', APIkey))
  
  
  # Pull out the data and convert it to a tibble
  pumstbl <- as_tibble(fromJSON(rawToChar(pums$content)))
  colnames(pumstbl) <- c(pumstbl[1,])
  pumstbl <- slice(pumstbl, -1)
  pumstbl
  
  
  # Convert variables to the correct type
  # Numeric (PWGTP, AGEP, GASP, GRPIP, JWAP, JWDP, and JWMNP)
  for (i in 1:ncol(pumstbl)) {
    if (colnames(pumstbl)[i] %in% allowNumVar) {
      pumstbl[,i] <- as.numeric(unlist(pumstbl[,i])) 
    }
  }
  
  # Time - replace the numeric code value with the time (JWAP, JWDP)
  if ('JWAP' %in% colnames(pumstbl)) {
    pumstbl <- pumstbl |>
      mutate(JWAP = recode_values(JWAP, 
                                  from = jwapValues$item, 
                                  to = jwapValues$midpoint))
  }
  if ('JWAP' %in% colnames(pumstbl)) {
    pumstbl <- pumstbl |>
      mutate(JWDP = recode_values(JWDP, 
                                  from = jwdpValues$item, 
                                  to = jwdpValues$midpoint))
  }
  
  # Categorical - convert to factor (FER, HHL, SCH, SCHL, and SEX)
  if ('FER' %in% colnames(pumstbl)) {
    pumstbl <- pumstbl |>
      mutate(FER = factor(FER, 
                      levels = c('0', '1', '2'), 
                      labels = c('N/A (less than 15 years/greater than 50 years/ male)', 'Yes', 'No')))
  }
  if ('HHL' %in% colnames(pumstbl)) {
    pumstbl <- pumstbl |>
      mutate(HHL = factor(HHL, 
                      levels = c('0', '1', '2', '3', '4', '5'), 
                      labels = c('N/A (GQ/vacant)', 'English Only', 'Spanish', 'Other Indo-European languages', 'Asian and Pacific Island languages', 'Other Language')))
  }
  if ('SCH' %in% colnames(pumstbl)) {
    pumstbl <- pumstbl |>
      mutate(SCH = factor(SCH, 
                      levels = c('0', '1', '2', '3'), 
                      labels = c('N/A (less than 3 years old)', 'No, has not attended in the last 3 months', 'Yes, public school or public college', 'Yes, private school or college or home school')))
  }
  if ('SCHL' %in% colnames(pumstbl)) {
    pumstbl <- pumstbl |>
      mutate(SCHL = factor(SCHL, 
                      levels = c(schlLevels), 
                      labels = c(schlLabels)))
  }
  if ('SEX' %in% colnames(pumstbl)) {
    pumstbl <- pumstbl |>
      mutate(SEX = factor(SEX, levels = c('1', '2'), labels = c('Male', 'Female')))
  }
  

  # Replace numbers indicating NA with NA in the numeric variables, if they have an NA value (GASP, GRPIP, and JWMNP)
  if ('GASP' %in% colnames(pumstbl)) {
    pumstbl$GASP[pumstbl$GASP == 3] = NA
  }
  if ('GRPIP' %in% colnames(pumstbl)) {
    pumstbl$GRPIP[pumstbl$GRPIP == 0] = NA
  }
  if ('JWMNP' %in% colnames(pumstbl)) {
    pumstbl$GRPIP[pumstbl$GRPIP == 0] = NA
  }
  
  
  # Change class of tibble
  class(pumstbl) <- c('census', class(pumstbl))
  
  
  # Return data
  return(pumstbl)
  
}
```


```{r}
# Convert the 'bRange' and 'eRange' variables into time objects
  jwapValues$period <- as.character(hm(jwapValues$eRange)-hm(jwapValues$bRange))
  jwapValues$midpoint <- hm(jwapValues$bRange) + as.period((as.duration(hm(jwapValues$eRange)-hm(jwapValues$bRange))/2))
  #Doesn't work because a.m. and p.m. - need them to be AM and PM
  
  # Replace 'a.m.' and 'p.m.' with 'AM' and 'PM' for easier parsing to time class
  jwapValues$bRange <- str_replace_all(jwapValues$bRange, c('a.m.' = 'AM', 'p.m.' = 'PM'))
  jwapValues$eRange <- str_replace_all(jwapValues$eRange, c('a.m.' = 'AM', 'p.m.' = 'PM'))
  # Convert the 'bRange' and 'eRange' variables into time objects
  jwapValues$bRange <- parse_date_time(jwapValues$bRange, '%I:%M %p')
  jwapValues$eRange <- parse_date_time(jwapValues$eRange, '%I:%M %p')
  jwapValues$midpoint <- jwapValues$eRange - jwapValues$bRange
  # Actually don't need to do all these because three of the time periods actually are 9 minutes instead of 4 which is stupid and annoying, actually I do need it because it's the only way Ive found to deal with AM/PM
  
    # Calculate the middle of each time period
  jwapValues$midpoint <- hm(jwapValues$bRange) + as.period((as.duration(hm(jwapValues$eRange)-hm(jwapValues$bRange))/2))
  # doesn't work because it ends up with a period instead of a time
  
```


Values to test function with
year <- '2024'
numericVariables <- c('PWGTP', 'AGEP', 'GASP', 'GRPIP', 'JWAP', 'JWDP', 'JWMNP')
categoricalVariables <- c('FER', 'HHL', 'SCH', 'SCHL', 'SEX')
geographyLevel <- 'STATE'
specificGeography <- 'Maine/ME'
APIkey <- '533d84cb67d9d71606e0baf0f4e7f32e4af7c4ce'

```{r}
pumsTest <- GET('https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=533d84cb67d9d71606e0baf0f4e7f32e4af7c4ce')

pumsTest2 <- GET('https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=533d84cb67d9d71606e0baf0f4e7f32e4af7c4ce')
pumstbltest2 <- as_tibble(fromJSON(rawToChar(pumsTest2$content)))

pumsTest4 <- GET('https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&for=STATE:16&SCHL=24&key=533d84cb67d9d71606e0baf0f4e7f32e4af7c4ce')
pumstbltest4 <- as_tibble(fromJSON(rawToChar(pumsTest3$content)))
```

```{r}
 pumstblOg <- as_tibble(fromJSON(rawToChar(pums$content)))
  colnames(pumstblOg) <- c(pumstblOg[1,])
  pumstblOg <- slice(pumstblOg, -1)
  pumstblOg
```

