---
title: "Riley_Project1"
format: html
---

# Project Description
This project will manipulate and process data sets from a census API and then summarize and plot returned data, mainly using custom functions.

# Author Information
Author: Shannon Riley
Organization: NCSU ST 558
Date: Fall 2026 (Completed ___)

## Data
This project will utilize the Public Use Microdata Sample (PUMS) Census API. We will consider the years 2021-2024 for the basic 1-Year Estimates (will not include Puerto Rico).

```{r}
#| echo: false
#| output: false
# Load required packages
library(dplyr)    # to perform basic EDA
library(ggplot2)  # to make nice plots
library(httr)     # to query an API
library(jsonlite) # to parse results of API query, tidyjson is another possible package to use
library(readr)    # to read in CSV data
library(stringr)  # to create clean column names
library(tidyr)    # to clean up tibbles
```

First, the JSON files with the variable descriptions for each year (2021-2024) will be downloaded. We only want to parse the following subset of variables:
+ AGEP - Age
+ GASP - Gas cost (monthly cost, use ADJHSG to adjust GASP to constant dollars)
+ GRPIP - Gross rent as a percentage of household income past 12 months
+ JWAP - Time of arrival at work (hour and minute)
+ JWDP - Time of departure for work (hour and minute)
+ JWMNP - Travel time to work
+ PWGTP - PUMS person weight
+ FER - Gave birth to child within the past 12 months
+ HHL - Household language
+ SCH - School enrollment
+ SCHL - Educational attainment
+ SEX - Sex
+ ST (for 2021, 2022)/STATE (for 2023, 2024) - State Code based on 2020 Census definitions
+ REGION - Region code based on 2020 Census definitions
+ DIVISION - Division code based on 2020 Census definitions

Pull each year's list variable values using the *GET()* function from the *httr* package:
```{r}
varVal2021 <- GET('https://api.census.gov/data/2021/acs/acs1/pums/variables.json')
str(varVal2021, max.level = 1)
```